### 2023-03-02 calculus.tex
Shorten intro of section 2

% TODO elaborate on this part

Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
such ill-typed ones meaningless. This is insufficient for editor services and language servers,
which must provide users feedback on where and why a program is ill-typed. Moreover, they must
continue providing semantic services even when an error is encountered. Indeed, similarly to
error-recovering parsers, real-world type checkers must robustly recover from type errors.
Traditionally, this has been performed in an \emph{ad hoc} manner [citations?], with different
languages and tools adopting varying approaches in the localization, reportage, and recovery of
errors.

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
bidirectional type error localization. Namely, the calculus employs bidirectional and gradual typing
to provide a judgmental framework with a unified theory for the design of typing semantics that
support ill-typed programs. \Cref{sec:calculus-calculus} describes its form and metatheory by
example on a small gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary
product types in \cref{sec:calculus-products}. All rules and theorems may be found in the
supplementary appendix, alongside a complete mechanization in the Agda proof assistant
\cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With products, we then consider how
error localization is complicated by patterned let expressions in \cref{sec:calculus-let}. Finally,
\cref{sec:calculus-poly} explores how the system might be extended to more complex typing features,
such as System F-style polymorphism.

% TODO not really sure when to start talking about bidirectional typing

% \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
% type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
% system in which type information is recursively propagated through terms and allows for local type
% inference.

% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
% bidirectional?

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
bidirectional type error localization and recovery. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
the system might be extended to more complex typing features, such as System F-style polymorphism.
